\documentclass[10pt,a4paper,usenames,dvipnames]{article}
\usepackage{amssymb,amsmath}
\author{ Jan Groothuijse }
\title{ Proving an expression compiler correct }
\begin{document}

\maketitle
\begin{abstract}
This document is the report accompanying a coq file in which a simple expression compiler is proven correct. 
It shall discuss the challenges and solutions of the implementation, focussing on the process of creating it and design decisions.
For further (non process or design related) documentation please see the accompanying coq file. Since this subject was number
3 on a list of possible subjects and to ease searching in the coq file, all assignments are prefixed by '3.'. 
Those assignments also form the structure of this document.
\end{abstract}

\section*{3.1 Inductive definition of \emph{Exp}}
To keep the definitions small and aid the provability of the theorems, I decided to have a seperate \emph{BinOp} definition, so that the actual \emph{Exp} definition only needed 2 clauses. Since I have not tried it the other way, I do not know if this was the right decision. When proving the correctness of the compiler I did need to do induction on \emph{BinOp} in order to keep the goals readable and I did end up with what I think is not the shortest way to prove the theorem.

For the definition of the following functions, these 'small' datatypes did work well.

\begin{align*}
& \text{Inductive} \: BinOp := \\
& \vert \: add : BinOp \\
& \vert \: sub : BinOp \\
& \vert \: mul : BinOp \\
\\
& \text{Inductive} \: Exp :=  \\
& | \: lit : nat \rightarrow Exp \\
& | \: exp : BinOp \rightarrow Exp \rightarrow Exp \rightarrow Exp. \\
\end{align*}

\section*{3.2 Eval }
Defining \emph{eval} based on the previously declared datatypes was pretty straightforward.

\section*{ Front-end }
I decided for testing purposes and for dusting off my ocalm/coq 'experience' to implement a front-end for this compiler.
Since I already build a simple toy compiler once using a functional language, the design was mostly in my head already.

I did at some point make a version that used mutual recursion, but in order to show coq that the arguments where indeed descreasing I had to duplicate some logic in all of the functions. So it turned out to be really big and I decided to use
this approach instead. This version however does not respect operator precedence.

Its a simple lexer-parser design, with the note that the lexer's result is reversed, a parser could be designed to handle this, but in this case it makes no difference.

Along the way I also created a \emph{option\_flatten}, which turns an \emph{option (option a)} into \emph{option a}. I did find a \emph{option\_map}, but I could not find a function that does what my option\_flatten does so I created it.

\section*{3.3 Inductive definition of RPN }

In the reverse polish notation, because there are no parentheses, there are only numbers and operators. While this could be defined using a \emph{list (either nat Binop)}, i decided to play it safe and define a \emph{RPNSymbol} to be either a literal (number) or an operator (\emph{BinOp}). I did however use lists, \emph{RPN} is defined as \emph{list RPNSymbol}.

\begin{align*}
& \text{Inductive} \: RPNSymbol := \\
& | \: rpnlit : nat \rightarrow RPNSymbol \\
& | \: rpnop : BinOp \rightarrow RPNSymbol. \\
\\
& \text{Definition} \: RPN := list \: RPNSymbol.
\end{align*}

The reason for not using \emph{Either} is readability of the proof, I was afraid to get lost by some many lefts and rights. The reason to use lists is because it has all sorts of functions, lemma's and theorems already defined for it.

\section*{3.4 Compiler }
The compiler is pretty straightforward, using lists as output did help here. Its the shape of the recursion that does the transformation.

\section*{3.5 Evaluator for RPN}
When evaluating rpn, one needs a 'stack' like datastructure. Fortunaly for us the list is such a datastructure, so our stack of numbers is a list nat. The \emph{rpn\_eval\_} function consumes code and uses the stack parameter as accumulator. In order to create a nicer interface I also created \emph{rpn\_eval} which only takes RPN as argument and then call \emph{rpn\_eval\_} with an empty stack. This should be the function that is 'exported'.

This wrapper function poses no real challanges for proving, but it should be unfolded before doing induction so that the induction hyptothesis is somewhat more usefull.

The definition of \emph{rpn\_eval\_} looks slightly clumbsy because in case of \emph{rpnop} we need to pop 2 values from the stack. The order of \emph{n1} and \emph{n2} is reversed to operate more like the semantics defining \emph{eval} function.

\section*{ Append properties }

For reasons unknown to me, \emph{app\_nil} and \emph{app\_assoc} where not defined on my system even after \emph{Require Import List}. I used version 8.2pl2 (September 2010) of CoqIde. Because my definitions make use of appends i defined and proved my own \emph{app\_nil} and \emph{app\_assoc}.

\section*{ 3.6 Prove that $\forall$ e $\in$ Exp, Some (eval e) = rpn\_eval (rpn e) }

At some point when I was trying to prove the above statement directly it occured to me that I needed to use the intermediate steps produced by running \emph{rpn\_eval\_}. When doing induction on the structure of \emph{Exp}, it does not help to know that smaller expressions also return a result if we cannot use that result. In the intermediate step it would be easier to show what happens on the stack.

So I first proved a different theorem concentrating mostly on the inner workings of the \emph{rpn\_eval\_} function: Theorem step:
\begin{multline*}
	\forall e \in Exp, \forall s \in $\emph{list nat}$, \forall t \in RPN, \\
	rpn\_eval\_ s (app (rpn e) t) = rpn\_eval\_ (eval e :: s) t
\end{multline*}

Because this theorem does incorporate a stack and does not concern itself with returning the actual result, the induction hypothesis is more helpful and the theorem becomes easier to prove.

After proving the \emph{step} theorem the original theorem becomes a corollary.

\subsection*{ Proof normalization (or lack thereof) }
Parts of the proof of the \emph{step} theorem are highly repetetive, those parts also happen to be rather lenghty and as a result the entire proof is very long. At some point I use induction on the binary operators, if I could postpone doing so the proof could be much shorter. This way however the proof-goals stay very readable and understandable, which made my life slightly easier.

\section*{ 3.7 Variables }
To allow for variables in an environment, I chose refer to those variables by numbers and use a list to actually store their values. This meant that the expressions now have 2 additional constructors, one to get a value from memory and one to update a a value in memory.

\begin{align*}
& \text{Inductive} \: Exp2 := \\
& | \: lit2 : nat \rightarrow Exp2 \\ 
& | \: var : nat \rightarrow Exp2 \\
& | \: letvar : nat \rightarrow Exp2 \rightarrow Exp2 \rightarrow Exp2 \\
& | \: exp2 : BinOp \rightarrow Exp2 \rightarrow Exp2 \rightarrow Exp2 \\
\end{align*}

Now in the \emph{eval2} function we have the environment as extra argument and 2 additional cases. To implement scoping rules
in this function we use ordinary backtracking.

For the \emph{RPN} version I needed to define more symbols, one to get a value from memory and one to set a value. Because we cannot use backtracking here to implement scoping rules I needed 2 additional symbols, one to store a specific value and one to rollback this store action. These addaptations makes the symbols look slightly more like bytecode.

\begin{align*}
& \text{Inductive} \: RPNSymbol2 := \\
& | \: rpnlit2 : nat \rightarrow RPNSymbol2 \\
& | \: rpnvar : nat \rightarrow RPNSymbol2 \\
& | \: popframe : RPNSymbol2 \\
& | \: pushframe : RPNSymbol2 \\
& | \: rpnop2 : BinOp \rightarrow RPNSymbol2. \\
\\
& \text{Definition} \: RPN2 := list \: RPNSymbol2. 
\end{align*}

In the \emph{rpn2} function whenever a letvar is compiled, we first execute the computation of the value to be stored, then
put the variable number on the stack and then do a pushframe, this will case a new memory frame to be made with as value for the variable number the computation we have done earlier. With this frame pushed on top of the existing frames we execute the 'in' expression and after that expression we pop the frame before continuing, so that the memory operation has been undone.

The \emph{rpn2\_eval\_} function like the \emph{eval2} function takes an additional argument. But instead of one \emph{list} to represent memory, \emph{rpn2\_eval\_} uses a list of lists of nat to represent a stack of memory frames.

\section*{3.8 Avoiding None }



\end{document}